Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-06 paper.tex
removed long comments
state of the art updated

%\documentclass[a4paper, 10pt]{IEEEtran}      % for icra/RAL final
\documentclass[usletter, 10pt, conference]{ieeeconf} % initial submission to RAL/ICRA
\usepackage{graphicx}
\usepackage{booktabs}
%\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{subfig}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{multirow}
\usepackage{flushend}
\usepackage[bookmarks=false]{hyperref}
\usepackage{setspace} % for setstretch in algorithm
\pdfminorversion=4

@@

%D. Bedn\'{a}\v{r}:
%Faculty of Science, Masaryk University, Kamenice 5, 625 00 Brno, Czech Republic\\

@@

Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
The size and location of tunnels determine the ligand size and shape which can enter the active site.
%The tunnel has to be wide enough and also the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.
Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
This helps them to reduce the number of necessary in-vitro lab experiments.
The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
However, as the ligands are typically non-spherical, these methods can often skip possible candidate tunnels.

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{-0.5}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{ccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3}  \\
Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
             &            & (orange)         \\  %& trajectory
\multicolumn{3}{c}{%
\includegraphics[width=0.16\textwidth]{fig/renderDCP}  \hskip 15pt
\includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
%\hbox{
%\vbox{
%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
%\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
%} 
%}
\end{tabular}
}
\caption{\label{fig::motiv}
    Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
}
\end{figure}

@@

%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
%The knowledge about the existence of the tunnels and their properties (e.g., their bottleneck and chemical properties around the tunnel),
%is important to understand interactions between proteins and other molecules (ligands)~\cite{Koudelakova2013}.
%As the interaction takes place in the active site, the existence of a tunnel connecting the active site with the outer environment
%is necessary for the interaction.
%To get the ligand to the active site, a transportation tunnel connecting the protein outer environment with the active site has to exists.
%Biochemists decide if a tunnel can be used to transport a given ligand mainly based on the tunnel length and bottleneck (i.e., the radius of the smallest sphere that forms the tunnel).
%This is used for example in protein engineering, where the task is to change selected properties of a protein, e.g., its stability under different outer conditions~\cite{Koudelakova2013}. % or its activity of the protein towards other molecules~\cite{Pavlova2009}.
%The design of suitable ligands causing the desired changes can be speeded up by detecting and analyzing the tunnels leading to the active sites.
%This can be achieved by detecting and studying so called tunnels in proteins which can serve as the transportation paths for the 
%ligand from the outside environment to the active site or vice versa. 
%Whereas the tunnel computation is already a well established research field, the simulation of ligand transportation through the detected tunnels is rather new.
%Ligands are typically of non-spherical shape, therefore it is difficult to estimate their traversability through tunnels computed for a spherical probe.
%The decision based only on spherical tunnels requires previous expertise in the domain and yet it may be imprecise.

%Proteins are dense structures which limits movements of the ligand insider the tunnels, which leads to the narrow passage problem.
%To cope with this problem, we propose to utilize the principle of guided sampling~\cite{vonasek2009rrt,denny2016dynamic}.
%It helps to keep also potential solutions which do not fit to the geometric restrains but can be still feasible because of the physico-chemical properties.
%This strategy helps to overcome the limitations of the conformational discretation introduced by the rotamer approximation.

%The proposed method can be seen as an extension for the traditionally used tunnel detection tools.
%Besides trajectory computation, visualization of the results in important. %we also propose methods for their visualization.
%Our motivation is to help the biochemists to perform so called virtual screening, where they test the traversability of a ligand through a given tunnel.
%In order to predict the success of ligand traversability, the virtual screening performs hundreds of thousands of tests and checks if the ligand passed through the tunnel.
%The advantage of the proposed approach is that we can search the ligand path inside a specific tunnel which substantially decreases the computational time and resources required for the virtual screening.
%We further propose visualization techniques for presenting the results to biochemists.
%Therefore, methods for visualization of the results  are discussed.

@@

%The analysis of protein structure aiming to reveal the tunnels has been supported by different computational software tools which take the geometry of the protein as an input and explore the inner void space (e.g., CAVER 1.0~\cite{petrek2006caver} or MOLE~\cite{Petrek20071357}). 
%Early methods for tunnel detection utilized a discretized 3D grid, where each cell is considered as occupied or free depending
%on the presence of atoms of the protein~\cite{petrek2006caver,Petrek20071357}.

%TODO some words about MD and why it better to analyze stuff with tunnels than with MD

@@

%This can be formulated as a motion planning problem in a high-dimensional configuration space $\C$.
%In this case, the configuration space $\C$ has at least $6+n$ dimensions (3D rotation + 3D translation + additional degrees of freedom
%caused by the ligand flexibility).

%Tunnels can be then searched using standard graph-search methods, like Dijkstra's algorithm.
%Besides, the grid can be used to identify other relevant properties like 
%pockets, cavities, or channels~\cite{sehnal2013mole,petrek2006caver}.
%One of the first grid-based approaches to the detection of tunnels in protein is the CAVER 1.0 algorithm by Pet\v{r}ek et al.~\cite{citeulike:6257975}.
%The obvious disadvantage of the grid-based methods is the high memory demand and their dependency on the grid resolution.
%Due to the high memory consumption, these methods are not suitable for tunnel detection in dynamic proteins and therefore they
%are used primarily for analysis of static molecules or individual snapshots of molecular dynamics.
%Currently the most widely used approach to tunnel detection is based on ordinary Voronoi diagrams (VD) or Weighted Voronoi Diagrams (WVD). %~\cite{caver3,yaffe2008}.
%The ordinary VD is computed on points representing centers of all atoms, without considering the radii of atoms.
%This may lead to detection of tunnels with incorrect bottlenecks. % i.e., incorrect radius of a smallest probe that can traverse the tunnels.
%To consider atoms with different radii, the weights of individual points are determined by the van der Walls radii of atoms in WVD.
%An alternative solution is to compute a non-weighted VD on an extended point set, where 
%each atom is approximated by several spheres with a small radius~\cite{yaffe2008,caver3}.
%VD-based methods are memory less demanding and also faster than the grid-based methods.
%The extension of VD-based methods to dynamic molecules requires to construct VD in the frames being analyzed and 
%finding correspondences between them.
%The existing approaches often use hierarchical clustering to match Voronoi vertices and edges from different frames, which is computationally demanding~\cite{lindow2012dynamic,caverDetails}.

%The protein tunnels computed for spherical probes contain valuable information about the protein void space.
%Based on the biochemical properties, the chemists can decide if a given tunnel is relevant and can be possibly used by a ligand to exit the protein.
%The biochemically relevant tunnels can be then analyzed separately.

%The configuration space $\C$ is formed by all possible configurations of the ligand in the tunnel, i.e., considering its rotation, translation, and possibly also other degrees of freedom responsible for the conformation changes.
%The dimension of the configuration space is given by the degrees of freedom (DOF) of the ligand, i.e., 6D for a rigid ligand and 6D + $n$ for a flexible
%ligand with $n$ DOFs.
%Sampling-based motion planning methods can be used to search this high-dimensional configuration space~\cite{Lav06}.

%Sampling-based motion planning methods are suitable for computing the trajectories of the ligand as they 
%can cope with many-DOF robots (objects) of arbitrary shapes.
%The flexibility of ligands (or even proteins) can be modeled using a multi-link kinematic chain, where torsional angles
%can change~\cite{songPFpath}.
%This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid} or analysis of loop motions~\cite{cortes2004geometric}.
%This is useful, e.g., in the protein folding studies~\cite{al2012motion,gipson2012computational,amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro} or analysis of loop motions~\cite{cortes2004geometric}.
%Sampling-based methods have also been used for 
% tunnel detection~\cite{vonasek2016application,vonasek2017tunnel} and
% exit pathway computation~\cite{cortes2010simulating,guieysse2008structure}.
%none of the existing work focuses on the trajectory generation in tunnels.
%However, not all conformations are feasible, which has to be checked using potential energy, which is time consuming.
%The idea of sampling-based motion planning is to randomly sample the configuration space $\C$ and classify the samples as free or non-free using collision detection.
%The free samples are stored in a roadmap (a graph structure), in which a path can be searched using standard graph-search methods.

Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
incrementally builds a configuration tree rooted at the initial configuration.
In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
If $\qnew$ is collision-free, it is added to the tree.
The algorithm terminates if the tree approaches the goal configuration close enough.

%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.

%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
%Sampling-based planners can provide poor performance in the presence of narrow passages, as many iterations
%are required in order to sample the narrow passages dense enough.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in given regions.
These regions can be estimated using workspace information, such as by medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}, 
or they can be identified, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
growth of the configuration tree towards the random samples may be blocked by obstacles~\cite{vonasekphd}.
To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
The radius of the ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
The disadvantage of DD-RRT is the sensitivity to the predefined radius.
Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying
the method to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path is used
to generate samples in the configuration space.
The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}. %\cite{he2016efficient} in RAL!! + zhang2008fast
Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
the retractions only on the narrow passages.
Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis in order to improve the growth in the narrow passages.
In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed towards the medial axis.
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.

%TODO NEKAM PRESUNU/SMAZU: 
%A proper distance metric in the configuration space is necessary in order to determine which samples should be connected together (in the case of PRM), or which node of the tree should be expanded (in the case of RRT).
%Generally, it is not easy to combine the rotational and transnational components of the configuration space.
%Moreover, the selection of the metric is also influence by the shape of the objects (robots).
%The flexible ligands can be modeled as a multi-link kinematic chain~\cite{songPFpath,cortes2005path}, but it significantly
%increases the number of DOFs and consequently the dimension of the configuration space.
%Moreover, by considering flexible ligands, their shape differ in each conformation, which makes the selection of a suitable metric difficult.

@@

%from wiki:
%Multiple static structures experimentally determined for the same protein in different conformations are often used to emulate receptor flexibility.[19] Alternatively rotamer libraries of amino acid side chains that surround the binding cavity may be searched to generate alternate but energetically reasonable protein conformations.[20][21]
% 19 = \cite{totrov2008flexible}
% 20 = \cite{hartmann2009docking}
% 21 = \cite{taylor2003fds}

@@

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
%The set of all feasible configurations at scale $s$ is denoted $\CFD \subseteq \C$.

@@

%While the nearest neighbors are searched using 6D Euclidean metric, which can be efficiently realized using KD-trees, the expansion
%procedure utilizes a metric that is sensitive to the shape of the ligands.
%This is necessary as the ligands adopt various shapes according to the actual conformations.
%To minimize the distance between $\qrand$ and the ligand, the shape of the ligand should be considered.

%Therefore, we propose to measure the distance as the smallest 3D Euclidean distance between atoms placed at $\qrand$ and a configuration $q$ being tested.
The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
is also the nearest point of the ligand at $q$ to $\qrand$.

\begin{algorithm}[th]
{\small
\setstretch{0.88}
\caption{\label{alg::expand}expand}
\KwIn{
   tree $\T$,
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$,
}
%\KwOut{
%    set $S$ of collision-free configurations reachable from $\qnear$\;
%}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \nllabel{alg::expb}
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to the next conformation
        }
    }
}
}
\end{algorithm}

@@

\begin{figure}[hb]
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

%\subsection{Visualization of Results}
%
%TODO Bara: prepsat/motivace/jak se visuazlizuje
%toto je cast textu z toho posledniho paperu. Kdyztak to vyhod.
%
%%In order to fursupport the tunnel exploration process performed by biochemists.
%The above defined characteristics can be shown even as a graph, or better, presented visually by mapping them using colors to the tunnel spheres.
% Alternatively, the surface representation can be used to visualize the tunnel.
%In this case, a 3D point on the tunnel surface is colored according to the property value in its nearest tunnel part $i$,  i.e., such part $i$ whose center $c_i$ is the closest to the point among all tunnel parts.
%
%The examples of the color mapping are shown in Fig.~\ref{fig:properties}.
%The accessibility (Fig.~\ref{fig:properties}a) shows that more than the half of the tunnel is not accessible (red part of the tunnel).
%The throughput shows (Fig.~\ref{fig:properties}b) that the only difficult is the part around the bottleneck (red color in Fig.~\ref{fig:properties}b), and the second half of the tunnel is also traversable.
%
%\begin{figure*}[h]
%\centering
%\begin{tabular}{ccc}
%\includegraphics[width=0.3\textwidth]{fig/accessibility} &
%\includegraphics[width=0.3\textwidth]{fig/throughput} &
%\includegraphics[width=0.3\textwidth]{fig/ligand-scale} \\
%  (a) & (b) & (c) \\                     
%\end{tabular}
%\caption{Tunnel properties mapped onto its surface. The tunnel begins at the top left corner.
%(a) Accessibility --- green denotes the parts that were accessed by the ligand easily ($A(i) > 80$~\%), 
%    while red denotes parts that were accessible  in $<10$~\% of cases.
%(b) Throughput --- green denotes parts that are highly probable to be passed by the ligand ($T(i) > 80$~\%), 
%    while red denotes those that were hard to pass ($T(i) < 20$~\%).
%(c) Ligand scale --- the average scale of the ligand was \tylde 50\% in orange parts, \tylde 60\% in yellow parts, \tylde 80\% in yellow to green parts and \tylde 100\% in green parts.
%\label{fig:properties}
%}
%\end{figure*}
%
%

%\def\cstart{\overline{c_{start}}}
%\def\cend{\overline{c_{end}}}

@@

%
%{\def\a{270}
%\begin{figure}
%\centering
%\renewcommand{\tabcolsep}{0pt}
%\begin{tabular}{cccc}
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-cdCount} &
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m001-oneAtom-rrt_path_time} &
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-cdCount} & 
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m037t-rrt_path_time} \\
%\multicolumn{2}{c}{One-atom} &
%\multicolumn{2}{c}{m037}  \\
%\multicolumn{4}{c}{%
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-cdCount} 
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-rrt_path_time} 
%\includegraphics[width=0.15\textwidth,angle=\a]{fig/m040-10-trajDistToGoal}} \\
%\multicolumn{4}{c}{m040} 
%\end{tabular}
%\end{figure}
%}
%

>>>>

\documentclass[usletter, 10pt, conference]{ieeeconf} % initial submission to RAL/ICRA
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{subfig}
\usepackage[ruled,lined]{algorithm2e}
\usepackage{multirow}
\usepackage{flushend}
\usepackage[bookmarks=false]{hyperref}
\usepackage{setspace} % for setstretch in algorithm
\pdfminorversion=4

@@

@@

Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
The size and location of tunnels determine the ligand size and shape which can enter the active site.
Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
This helps them to reduce the number of necessary in-vitro lab experiments.
The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
However, as the ligands are typically non-spherical, these methods can often skip possible candidate tunnels.

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{-0.5}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{ccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3}  \\
Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
             &            & (orange)         \\  %& trajectory
\multicolumn{3}{c}{%
\includegraphics[width=0.16\textwidth]{fig/renderDCP}  \hskip 15pt
\includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
\end{tabular}
}
\caption{\label{fig::motiv}
    Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
}
\end{figure}

@@

@@

@@

Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is a single-query sampling-based motion planning method that 
incrementally builds a configuration tree rooted at the initial configuration.
Sampling-based planners can cope with many-DOF systems and they have been successfully employed also beoynd robotics, e.g. to
identify binding sites~\cite{bayazit2001ligand} or to determine ligand exit pathway~\cite{cortes2010simulating}.
%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.

%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g.
by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identifed using watershed labeling.
by indetifying them in a grid-based represnetation of the workspace~
or they can be identified, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

RRT-based planners cannot cope with the narrow passage problem simply by increasing the probability of sampling in them, as the
growth of the configuration tree towards the random samples may be blocked by obstacles. %~\cite{vonasekphd}.
To prevent this blocking, DD-RRT~\cite{yershovaDDRRT} limits the selection of nodes for expansion to a small ball. 
The radius of the ball is set to infinity for new nodes, and decreases to a predefined radius if the node cannot be successfully expanded.
The disadvantage of DD-RRT is the sensitivity to the predefined radius.
Therefore, the authors of ADD-RRT~\cite{jailletADRRT} proposed to adapt the ball radius according to the success rate of the expansion.
To attract the tree towards a given region, random samples have to be generated there only if the tree can expand to this region.
In~\cite{kardossRRTKK}, the probability of sampling is increased in several waypoints close to narrow passages, but without specifying
the method to find these waypoints.
The generalization of~\cite{kardossRRTKK} is the guided sampling, where a whole path is used
to generate samples in the configuration space.
The path can be computed in the workspace~\cite{vonasek2009rrt}, or iteratively refined in the configuration space based on the solution of a relaxed version of the problem~\cite{bayazitIRC}.

@@

%Another techniques to sample the narrow passages dense enough is to retract the random samples towards regions that are believed
%to be located in narrow passages~\cite{zhangRetraction,lee2012srrrt}.
%the random sample $\qrand$ is connected to the tree if the line segment  from $\qnear$ to $\qrand$ is collision-free. Otherwise, the retraction-step is performed.
%The task of the retraction step is to find a contact configuration around $\qnear$ that minimizes the distance to $\qrand$.
In Retraction-based RRT~\cite{zhangRetraction}, the tree is retracted along the boundary of the obstacles in the configuration space.
This requires the analysis of the contact space, which leads to the generalized penetration depth computation of high complexity~\cite{he2016efficient}. %\cite{he2016efficient} in RAL!! + zhang2008fast
Selective Retraction-based RRT~\cite{lee2012srrrt} avoids the expensive growth in the open areas of the configuration space and focuses 
the retractions only on the narrow passages.
Obstacle-based RRT~\cite{amatoOBRRT} utilizes several expansion procedures based on, e.g., random vectors, obstacle vectors, or medial axis.
In MARRT~\cite{denny2014marrt}, the newly generated samples are pushed towards the medial axis.
Difficult regions can be also identified manually by a human operator~\cite{denny2018general}. 
%The operator can also select a suitable planner for the region.<F12>
%Therefore, many algorithms uses a heuristic to compute the samples near boundary of the configuration space~\cite{amatoOBPRM,amatoOBRRT}.

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using Probabilistic Roadmaps.
Each configuration in the roadmap is also evaluated using electorstatic energy.
The high-energy configurations are discared, while low-energy conformations (and their vicinity) are sampled more dense.
The approach~\cite{singhLIG} can be also used to predict ligand binding sites.

%From reviews: vojta will solve
%\cite{hsuBridge} uz tam je
%\cite{bergWIG}
%\cite{denny2018general}
%\cite{singhLIG}
%\cite{bayazit2001ligand} - it's in the text already.

@@

@@

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.

@@

The atom-based metric can be evaluated fast using KD-trees with complexity $\mathcal{O}(\log n)$, where $n$ is the number of atoms of the ligand.
For each conformation $\l \in \L$ of the ligand, a KD-tree is built for the ligand at the configuration $q=(0,0,0,0,0,0,l,\smax)$
(i.e., for ligand with no translation and no rotation).
To find the nearest atom to $\qrand$ for a ligand at the configuration $q=(x,y,z,r_x,r_y,r_z,l,s)$, the translation vector $V=(x,y,z)$ and
rotation matrix $R$ given by $r_x$, $r_y$ and $r_z$ are computed.
Then, $\qrand$ is translated to $\qrand' = R^{-1}(\qrand-V)$ and the KD-tree search is used to find the nearest point towards $\qrand'$, which
is also the nearest point of the ligand at $q$ to $\qrand$.

\begin{algorithm}[th]
{\small
\setstretch{0.88}
\caption{\label{alg::expand}expand}
\KwIn{
   tree $\T$,
   configuration $\qnear$ to be expanded,
   random configuration $\qrand$
}
\KwData{
   ligand conformations $\L$,
   scale limits $\smin, \smax$, and $\sdelta$,
}
\hrule
\ForEach{$l \in \L$}{
    \ForEach{$s \in (\smax,\smax-\sdelta, \ldots, \smin+\sdelta, \smin)$}{
        $\qnew = \emptyset$; // empty configuration \nllabel{alg::expa} \\
        \For{$i = 1,\ldots,m$}{
            $q=\qnear$\;
            $q.position$ = random 3D position around $\qnear$\;
            $q.rotation$ = random 3D rotation\;
            $q.l = l$\; 
            $q.s = s$\;
            \If{isCollisionFree($q$)}{
                \If{$\qnew = \emptyset$ {\bf or} $\da(q, \qrand) < \da(\qnew,\qrand)$}{ \nllabel{alg::expand:a}
                    $\qnew = q$\;
                }
            } 
        }
        \nllabel{alg::expb}
        \If{$\qnew \ne \emptyset$} {
            $\T$.addNode($\qnew$)\;
            $\T$.addEdge($\qnear,\qnew$)\;
            {\bf break;} // go to the next conformation
        }
    }
}
}
\end{algorithm}

@@

\begin{figure}[hb]
\centering
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

@@
